\section{Metatheory}
\label{sec:metatheory}

This section summarizes the main metatheoretical results for \thecalculus. All
definitions from the previous sections have been formalized in
Agda. We mark the theorems we proved mechanically with a little Agda bird after
the theorem name. Non-mechanized proofs are shown in the supplement.

\subsection{Type Safety for Declarative Typing}

Preservation
yields the usual safety invariant across reduction steps.

\begin{theorem}[Expression preservation\;\AgdaBird]
  If\/ $\HasType \SessCtx \E \T$ and $\E \EReduce \E*$, then
  $\HasType \SessCtx {\E*} \T$.
\end{theorem}

\begin{theorem}[Process preservation]
  If\/ $\ProcHasType \SessCtx \Proc$ and $\Proc \PReduce \Proc*$, then
  $\ProcHasType \SessCtx {\Proc*}$.
\end{theorem}

We complement preservation with progress. Expressions are not guaranteed to
reduce to values but may be blocked on a constant applied to a value. The
reduction of constants is part of the process semantics.

\begin{theorem}[Expression progress\;\AgdaBird]
  \label{thm:expression-progress}
  If\/ $\HasType \SessCtx \E \T$, then either $\E$ is a value, or there exist
  an evaluation context $\EC$, constant $\C$ and value $\V$ such that
  $\E=\ECapp{\EApp\C\V}$, or there exists an $\E*$ such that $\E \EReduce \E*$.
\end{theorem}

A process is concurrently blocked, if the active threads are blocked and no
channel pair can synchronize. The formal blocked-process
predicate is given in \cref{fig:blocked-expr-proc}.
Process progress guarantees reduction only for non-blocked
configurations. Thus, deadlocks remain possible.

\begin{theorem}[Process progress]
  \label{thm:process-progress}
  If\/ $\ProcHasType \SessCtx \Proc$, then either $\Proc \Congruent \PExp{\CUnit}$ or an expression in $\Proc$ is blocked on concurrent actions or
    there exists an $\Proc*$ such that $\Proc \PReduce \Proc*$.
\end{theorem}
The proof is deferred to \cref{sec:appendix-metatheory-ts}.

\subsection{Weak Simulation for Translation}

We relate source and target reduction through the process translation
$\TransP\Proc$, which maps a well-typed process $\ProcHasType \Ctx \Proc$ to its
run-time configuration under a value substitution $\Sub$ over the channel context
$\SessCtx$ of $\Ctx$. The translation introduces administrative
synchronization steps, hence both directions only hold up to a weak bisimilarity $\Bisim$,
which identifies run-time configurations that differ only by administrative
drop and acquire steps, channel discards, and the reordering of
independent binders.

\begin{theorem}[Forward simulation\;\AgdaBird]
  If $\Proc \PReduce \Proc*$\!, then there exists
  $\TProcQ$ such that $\TransP\Proc \TPReduce^{*} \TProcQ$ and
  $\TProcQ \Bisim \TransP{\Proc*}$.
\end{theorem}

The converse direction matches every target step by a finite source computation,
again up to $\Bisim$.

\begin{conjecture}[Backward simulation]
  For all $\TProcR$ and $\TProcQ$ with
  $\TProcR \Bisim \TransP\Proc$, if $\TProcR \TPReduce \TProcQ$, then there
  exists a source process $\Proc*$ such that $\Proc \PReduce^{*} \Proc*$ and
  $\TProcQ \Bisim \TransP{\Proc*}$.
\end{conjecture}
A proof sketch is provided in the supplement.


\subsection{Soundness and Completeness for Algorithmic Typing}

We now connect algorithmic typing with declarative typing. Soundness says
that any solved algorithmic judgment denotes a valid declarative typing, while
completeness says that declarative typings can be reconstructed algorithmically.

\begin{theorem}[Algorithmic typing soundness\;\AgdaBird]
  If either $\CheckType \E \AT \Eff$ or $\InferType \E \AT \Eff$
  and $\sigma \models \CSet$ is a closing
  substitution such that $\Fv(\AT) \subseteq \dom(\sigma)$, then
  $\HasType \Ctx \E {\AT{[\sigma]}}$.
\end{theorem}

% For completeness, we again first cover checking and then synthesis.
% 
% \begin{theorem}[Expression checking completeness]
%   If $\HasType \Ctx \E \T$, then
%   $\CheckType \E \AT \Eff$ for some type $\AT$, set of constraints $\CSet$,
%   and closing substitution $\sigma$ such that $\T = \AT{[\sigma]}$.
% \end{theorem}
% 
% \begin{theorem}[Expression inference completeness]
%   If $\HasType \Ctx \E \T$ and $\E$ is in synthesis form, then
%   $\InferType \E \AT \Eff$ for some type $\AT$, set of constraints $\CSet$,
%   and closing substitution $\sigma$ such that $\T = \AT{[\sigma]}$.
% \end{theorem}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../popl2027.tex"
%%% End:
