
\section{Related Work}
\label{sec:related-work}

\emph{Session Types.}
% \label{sec:session-types}
%
Context-free session types (CFSTs) have been introduced by
\citet{DBLP:conf/icfp/ThiemannV16} and further developed by
\citet{DBLP:journals/toplas/Padovani19} and
\citet{DBLP:journals/iandc/AlmeidaMTV22}. CFSTs liberate traditional 
session types from their tail-bias by replacing the prefix type
operators for the session operations by symmetric composition. This
move complicates the metatheory as the syntax of types is now
quotiented over the monoid laws and distributivity of composition over
choice. This complication makes type equivalence a decidable, but
difficult problem \cite{DBLP:conf/tacas/AlmeidaMV20} and subtyping an
undecidable problem
\cite{DBLP:journals/toplas/Padovani19,DBLP:journals/tcs/SilvaMV26}.

CFSTs are one starting point of our investigation. Symmetric
composition makes them well-suited to expressing the split for local
borrows. Only remote borrows require extra synchronization
machinery. Compared to the CFST work and the implemented system FreeST
based on it \cite{almeida26:_frees_concur_progr_languag_with}, our
calculus does not consider polymorphism. We rely on the FreeST
implementation as an external constraint solver for splitting
constraints.


The most closely related work is ``Borrowing from Session Types'' (BST) \cite{DBLP:journals/pacmpl/SaffrichS0V25}.
They present a similar approach to borrowing, but for regular session types, rather than CFST.
Our type system is heavily inspired by theirs, as it also relies on
BI-contexts that enable checking the correct run-time ordering of session operations.
We build on their translation from borrowing notation to explicit
splits, but adapt it to account for the distinction between local and
remote splits. The main technical differences are fourfold. First, BST
defines the semantics of borrowing by translation, whereas
\thecalculus has a direct operational semantics based on binder
groups. Second, \thecalculus distinguishes mobile, remote borrows from
local borrows (\cref{fig:bounded-session-types}); local
borrows are ordered by typing and evaluation and therefore avoid the
synchronization required by remote borrows. Third, splitting CFSTs
requires constraint generation modulo type equivalence rather than the
direct splitting algorithm available for traditional, tail-recursive
session types; we discharge these constraints with FreeST. Finally, we
define a low-level target calculus and prove a simulation result for
the translation, giving an implementation-oriented account of the
run-time synchronization protocol. The metatheory is mechanized in
Agda~\cite{Agda}.

Manifest sharing \cite{DBLP:journals/pacmpl/BalzerP17} is concerned with session types, which
make part of a protocol shareable. To access the shared part, they
employ a typed acquire-release pattern, which is reflected in the
types by different modalities. This approach makes the access pattern
manifest in the session type, similar to our pair of $\CDrop$-$\CAcq$
operators, but it may admit deadlock.  A subsequent work extends
manifest sharing to guarantee freedom of deadlock \cite{DBLP:conf/esop/BalzerTP19}.

Resource-passing style can also be hidden by monadic interfaces rather
than by borrowing. A number of session-typed APIs pursue this
direction~\cite{DBLP:conf/haskell/PucellaT08,DBLP:journals/scp/ImaiYY19,DBLP:conf/padl/NeubauerT04,DBLP:journals/jfp/Padovani17}.
\citet{DBLP:journals/jfp/Padovani17} builds a monadic layer over
session primitives that abstracts away explicit channel handling for
functions using a single session. This interface supports nested
sessions, but not interleavings among several sessions. Its primitive
operations decompose a session type, and monadic composition extends
this decomposition to prefixes of arbitrary length. Later systems
\cite{DBLP:conf/haskell/PucellaT08,sackman08:_session_types_haskel}
cover several sessions by representing the available capabilities as a
type-level stack. Their monadic composition decomposes prefixes of
multiple sessions at once, but the resulting prefixes remain coupled
rather than independent. \citet{DBLP:journals/corr/abs-1110-4163}
add session type inference to this line of work, and
\citet{DBLP:journals/scp/ImaiYY19} extend it with further
functionality. Compared with these approaches, our proposal avoids
type-level computation, permits interleaving of multiple sessions, and
splits distinct session types independently at arbitrary positions.

\emph{Ordered Types.}
% \label{sec:ordered-types}
%
Ordered types go beyond linear types in eliding the exchange
principle. Their formal basis is non-commutative logic
\cite{DBLP:journals/mlq/Abrusci90a, ABRUSCI199929} and they have been
introduced to a computer science setting by \citet{lambek58:_mathem_senten_struc,
  lambek60:_calcul_syntac_types} to describe
sentence structure in natural languages. Ordered lambda calculi
\cite{DBLP:conf/tlca/PolakowP99, DBLP:journals/entcs/PolakowP99}
bring the same non-exchange discipline to functional programming.

Early programming-language applications of ordered types and lambda
calculi are concerned with memory management and memory layout
\cite{DBLP:conf/popl/PetersenHCP03} or stack allocation
\cite{DBLP:conf/tldi/AhmedW03}. These works already illustrate a
pattern also followed by Law and Order, BST, and the present work:
ordered calculi are adapted to specific programming disciplines rather
than used in their pure proof-theoretic form. Each calculus picks up
the ordered structure, but adds substantial domain-specific elements,
to the extent that even basic elimination principles change.
\citet{DBLP:conf/tldi/AhmedW03} share with our work the use of
BI-contexts to distinguish ordered and unordered binding groups.

Closer to our setting, Law and order
\cite{DBLP:journals/pacmpl/Saffrich0024} is a precursor of BST. It
draws inspiration from non-commutative logic and ordered lambda
calculus and puts it to work in the context of typestate. It targets
sequential programming, so it is not concerned with synchronization
issues.

A different group of applications uses order to express consumption of
structured input. \citet{DBLP:journals/pacmpl/CutlerWNHGSP24} apply
ordered types to typing a calculus for streaming data. Ordering in
their system expresses temporal ordering in an input stream, that is, a
stream of type $s\cdot t$ is a stream of type $s$ followed by a stream
of type $t$. They rely on the same tree-shaped context structure we use, but with a
different interpretation: their parallel context former behaves
intuitionistically; contrary to ours, it allows for weakening and
contraction. This system is very close to the original theory of
ordered types, but it is restricted to first-order functions. This
line of work has predecessors that consider streaming of
tree-structured data using ordered linear types
\cite{DBLP:journals/jfp/KodamaSK08,DBLP:journals/jip/SatoSK11},
specialized to a fixed binary tree representation.

\citet{DBLP:journals/pacmpl/SchaeferVAN25} apply ordered linear
types to build a logical framework for the construction of verified
parsers. Order and linearity are used to describe the input string:
characters have to be processed in order and must not be duplicated by
the parser.

Finally, the subtyping rule
$(A;B) \| (C;D) \IsSSubtypeOf (A \| B) ; (C \| D)$ that weakens
parallelism for ordered sequentiality seems to
originate from work on concurrent Kleene algebra
\cite{DBLP:conf/concur/HoareMSW09}, but \citet{CAIRES2008120} mentions it
earlier in a type system for concurrent computation. It also appears
in work on behavioral separation \cite{DBLP:conf/popl/CairesS13}, attributed to
\citet{DBLP:journals/tcs/BloomE96a}.

\emph{Borrowing.}
% \label{sec:borrowing}
%
The notion of borrowing arose out of the need to control
aliasing in object oriented languages. It is closely tied to the notion of
ownership, which gave rise to numerous type disciplines to statically control
ownership. \citet{DBLP:series/lncs/ClarkeOSW13} present a survey of such
systems. Rust \cite{team24:_rust_progr_languag} is the most prominent
programming language supporting these concepts nowadays.
%
While ownership and borrowing are also useful concepts in the context
of session types, there are significant differences. Sessions are
resources for communication: using an endpoint advances its protocol
state. Thus there is no read-only analogue of an immutable borrowed
reference as it exists in Rust. Further, borrowing a session prefix
computes a residual session type, whereas Rust borrowing does not
temporally split ownership into a protocol prefix and suffix.

Other formalisms surrounding ownership and borrowing study
memory-like resources, including borrowing permissions
\cite{DBLP:conf/popl/NadenBAB12} and functional ownership systems
\cite{DBLP:journals/corr/abs-2310-18166}.
These systems address the interaction of linear ownership with
aliasing, permissions, and lifetimes. In our setting
aliases are endpoints for communication protocols: their use changes
the protocol state and, once sent to another process, may require
run-time synchronization.

\citet{DBLP:journals/pacmpl/WagnerGML025} derive borrowing from a
linear lambda calculus for manual memory management. Their Borrow
Calculus supports immutable and mutable borrows, lexical lifetimes, and
reborrowing without operational borrowing primitives; soundness is
semantic, via Borrow Logic, and yields termination and leak freedom.
This shares our view of borrowing as temporarily relaxing linear
resource passing, but our resources are session endpoints rather than
memory locations: each borrowed endpoint is a mutable communication
capability, borrowing computes a prefix/suffix decomposition, and
remote borrows require explicit synchronization. Thus our focus is on
context-free session type splitting, direct concurrent semantics, and
an implementation-level translation, rather than on a separation-logic
model of memory borrowing.

\citet{DBLP:journals/pacmpl/VillaniHDJ25} develop Tree Borrows, a
dynamic aliasing model for Rust that tracks derived references and raw
pointers in a tree of permissions. It likewise gives operational
content to borrows as derived capabilities whose use can restrict the
original owner and aliases, but only at the level of authority
tracking: Tree Borrows governs Rust pointer accesses and undefined
behavior, whereas our borrows split session types into a borrowed
prefix and a residual suffix. Remote borrows also require explicit
synchronization to order communication on the prefix before the suffix,
with no corresponding notion of protocol residual or cross-process
handoff in Tree Borrows.

Overall, prior borrowing systems manage aliasing of memory-like
resources, whereas our contribution is to make borrowing compute and
schedule portions of communication protocols.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../popl2027.tex"
%%% End:
