\section{Translation}
\label{sec:translation}

The calculus presented in \cref{sec:expr-proc,sec:types,sec:semantics}
records ownership and ordering in a linear type system and, dynamically, in
binder groups.  This section defines an untyped run-time calculus that is
closer to an actual implementation of splitting and borrowing.  The
run-time calculus provides the tools, ordinary channels
and synchronization cells, so that translated well-typed source processes
preserve the ordering guarantees established by the type system and the direct
semantics of \cref{sec:semantics}.

\begin{figure}
  \setlength{\abovedisplayskip}{0pt}%
  \setlength{\belowdisplayskip}{0pt}%
  \begin{align*}
    \TProc \grmeq
        & \TPExp\E
          \grmor \TPPar\TProc\TProc
          \grmor \TPNew[xy] \TProc
          \grmor \TPSync[\TEVarz]\Flag\TProc
      \tag{Configurations}
    \\
    \TCC
      \grmeq& \TCCEmpty
      \grmor \TPPar \TCC \TProc
      \grmor \TPNew[xy] \TCC
      \grmor \TPSync[\TEVarz]\Flag\TCC
      \tag{Configuration contexts}
    \\
    \Flag \grmeq & \FlagInit \grmor \FlagDropped %\grmor \FlagAcquired
      \tag{Synchronization flags}
  \end{align*}
  \caption{Untyped run-time syntax}
  \label{fig:unr-syntax}
\end{figure}

The syntax of constants, values, and expressions remains unchanged.
\Cref{fig:unr-syntax} defines target
configurations $\TProc$ and contexts $\TCC$.  The form $\TPNew[xy]\TProc$ is a
primitive-channel restriction with endpoints $\TEVar$ and $\TEVary$ scoped over $\TProc$; this is the
standard restriction construct of session calculi
\cite{Gay_Vasconcelos_2025}.  The form
$\TPSync[\TEVarz]\Flag\TProc$ introduces a synchronization cell $\TEVarz$ with state
$\Flag$ scoped over $\TProc$.  These cells are asynchronous
variables, similar to MVars
\cite{DBLP:journals/toplas/ArvindNP89,DBLP:conf/popl/JonesGF96}.  A
cell in state $\FlagInit$ records that the corresponding $\CDrop$ must happen next; a cell
in state $\FlagDropped$ records that the $\CDrop$ has happened and an $\CAcq$ may
proceed.

We call the channels of the target calculus \emph{primitive channels} to
distinguish them from (translated) source channels.  A
translated source channel is represented by a value
$\targettermcolor{c}=\TEChan[\V[1]]\TEVary[\V[2]]$, a triple consisting of an
optional pre-synchronization cell $\V[1]$, a primitive data channel
$\TEVary$, and an optional post-synchronization cell $\V[2]$.  The default
optional component is $\CUnit$.  A nontrivial pre-cell is consumed by
$\CAcq$ before the endpoint is used.  A nontrivial post-cell is signaled by
$\CDrop$ when ownership of the endpoint prefix is released.  Communication,
choice, and close operate on the primitive data channel; the extra cells encode
the ordering constraints on remote borrows that source binder groups
enforce directly. Local borrows do not use this machinery.

Target processes are quotiented by the congruence in
\cref{fig:translated-congruence}.  Its parallel and restriction laws are the
usual ones.  The additional laws commute independent synchronization binders
with each other and with channel binders, and extend synchronization scope
across parallel composition when the cell is fresh for the other component.
These laws let the reduction rules expose the primitive channel or
synchronization cell needed by a redex.

\begin{figure}
  \begin{mathpar}
    \RuleUProcRedExp      \and
    \RuleUProcRedDiscard  \and
    \RuleUProcRedFork     \and
    \RuleUProcRedNew      \and
    \RuleUProcRedLSplit   \and
    \RuleUProcRedRSplit   \and
    \RuleUProcRedDrop     \and
    \RuleUProcRedAcquire  \and
    %\RuleUProcRedCleanup  \and
    {\RuleUProcRedCom   \quad %\and
    \RuleUProcRedChoice}   \\
    \RuleUProcRedClose    \and
    \RuleUProcRedStruct   \and
    \RuleUProcRedPar      \and
    \RuleUProcRedBind     \and
    \RuleUProcRedSync
  \end{mathpar}
  \caption{Reduction of translated processes ($P' \TPReduce Q'$)}
  \label{fig:unr-process-reduction}
\end{figure}

The reduction behavior of translated expressions is the same as in the source
calculus (\cref{fig:expression-reduction}).  \Cref{fig:unr-process-reduction}
defines the target process reduction relation and uses $\ChanPat\E$ as an
abbreviation for triples $\TEChan[\U]\E[\V]$ when $\U$ and
$\V$ do not matter.

The target rules mirror the source process rules of
\cref{fig:process-reduction}, but make synchronization explicit.  Rule
\textsc{RU-New} introduces one primitive-channel restriction and two already-dropped
synchronization cells, corresponding to the leading acquire markers resulting
from source channel restriction.  Rule \textsc{RU-LSplit} performs a local split by
sharing the same primitive data channel between the two returned triples and
distributing the existing pre- and post-cells.  Rule \textsc{RU-RSplit}
allocates a fresh synchronization cell in state $\FlagInit$: the left channel receives it as a
post-cell, and the right channel receives it as a pre-cell.  Consequently
\textsc{RU-Drop} can signal the cell without blocking, while
\textsc{RU-Acquire} can proceed only after the signal is present.  The
communication, choice, and close rules synchronize on the primitive channel
component.  Remaining rules close reduction under parallel composition,
channel restriction, synchronization-cell restriction, and structural
congruence.

\begin{figure}
  \begin{mathpar}
    \RuleTransProcExp   \and
    \RuleTransProcPar   \and
    \RuleTransProcBind  \\
    \RuleTransBindSep   \and
    \RuleTransBindSeq   \and
    \raisebox{1.4\baselineskip}{%
      $\vcenter{\hbox{$\RuleTransBindVar$}\kern.8\baselineskip\hbox{$\RuleTransBindNil$}}$%
    }
  \end{mathpar}
  \smallskip
  \[
    \BindFlag{\Bind} = \FlagDropped \text{ if } \Bind = \BEmp;
    \quad \BindFlag{\Bind} = \FlagInit \text{ otherwise.}
  \]
  \caption{Translation of
    processes ($\Proc \TransProc \TProc*$)
    and binder groups ($\BindGroup \TransBind \TCC [\Sub]$). Function $\BindFlag{}$
    maps the segment immediately to the left of a separator to the initial
    state of its synchronization flag $\Flag$.}
  \label{fig:translation}
\end{figure}

\Cref{fig:translation} defines the translation relation.  It is parameterized
by a substitution $\Sub$ that maps source endpoint variables to translated
channel triples.  Rule \textsc{Tr-Exp} embeds a source process after applying
this accumulated substitution to its expression.  Parallel composition is
homomorphic.  A source restriction is translated by translating each binder
group against one endpoint of a fresh primitive-channel restriction, collecting the
substitutions produced by those binder groups, translating the body under the
extended substitution, and wrapping the resulting process in the
synchronization contexts produced by the binder-group translations.

The binder-group translation is the bridge between the direct semantics and
the target implementation.  Translating a variable records that the variable
denotes the current channel triple.  Translating a sequential group splits the
current triple locally: the first endpoint keeps the pre-cell, and the
remaining group keeps the post-cell.  Translating across a separator allocates
a synchronization cell.  The segment to the left receives the cell as a
post-cell, and the segment to the right receives it as a pre-cell.  The helper
$\BindFlag{\Bind}$ initializes the cell to $\FlagDropped$ when there is no
left segment, because no drop is pending; otherwise it initializes the cell to
$\FlagInit$.

The simulation result for this translation is stated in
\cref{sec:metatheory}.
Every reduction of a well-typed source process is matched by target reductions
of its translation.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../popl2027.tex"
%%% End:
