\input{popl2027-preamble}

% Build popl2027-supplement.tex first so main-text references to appendix
% labels are read from popl2027-supplement.aux. The aux file also contains
% package-generated labels, so clear those before the main aux file is read.
\externaldocument[][nocite]{popl2027-supplement}
\makeatletter
\newcommand*\IgnoreExternalLabel[1]{%
  \global\expandafter\let\csname r@#1\endcsname\@undefined}
\IgnoreExternalLabel{tocindent-1}
\IgnoreExternalLabel{tocindent0}
\IgnoreExternalLabel{tocindent1}
\IgnoreExternalLabel{tocindent2}
\IgnoreExternalLabel{tocindent3}
\IgnoreExternalLabel{tocindent4}
\IgnoreExternalLabel{tocindent5}
\IgnoreExternalLabel{TotPages}
\makeatother

\title{Context-Free Session Types with Borrowing}
\author{Janek Spaderna}
\email{spadernj@cs.uni-freiburg.de}
\orcid{0000-0002-9000-1239}

\author{Onur Sahin}
\email{sahino@cs.uni-freiburg.de}
\orcid{0009-0007-6908-0626}

\author{Peter Thiemann}
\email{thiemann@acm.org}
\orcid{0000-0002-9000-1239}

\author{Marius Weidner}
\email{weidner@cs.uni-freiburg.de}
\affiliation{%
  \institution{University of Freiburg}
  % \city{Freiburg}
  \country{Germany}}
\orcid{0009-0008-1152-165X}

% \authorrunning{J. Spaderna, O. Sahin, P. Thiemann, and M. Weidner}

% \institute{University of Freiburg, Freiburg, Germany}

%  Sentence 1: Modern [domain] systems increasingly rely on [technology], but they suffer from [general issue].
%  Sentence 2: Existing approaches fail to solve [specific bottleneck] due to [technical reason].
%  Sentence 3: In this paper, we introduce [System/Algorithm Name], a novel framework designed to [core mechanism].
%  Sentence 4: By utilizing [key optimization/technique], our approach reduces [metric] without sacrificing [other metric].
%  Sentence 5: Evaluation on [standard dataset/benchmark] shows that our method outperforms current state-of-the-art by [X]% in [metric].


\begin{document}
\begin{abstract}
Context-free session types (CFSTs) support expressive communication
protocols, but programming with them tends to expose the linear
threading of endpoints in both types and terms. This resource-passing
style makes modular session programs cumbersome, especially when a
component uses only a prefix of one or more protocols and returns the
residual endpoints to its caller.

We extend CFSTs with borrowing. A borrow splits a session type into a
borrowed protocol prefix and a residual suffix; for CFSTs, this split
must be computed modulo the equational theory of sequential
composition, choices, skipping behavior, and recursion. The resulting
calculus supports direct-style use of session endpoints while
preserving the linear invariant that communication advances protocol
state.

A key novelty is the distinction between local and remote
borrows. Local borrows remain in one process, where ordered linear typing
guarantees that the borrowed prefix is processed before the residual
suffix. Remote borrows may cross process boundaries; they require explicit
synchronization to ensure that the borrowed prefix is processed before
the residual suffix. The calculus has a direct operational
semantics, an algorithmic type system based on
constraints, and a translation to a lower-level calculus with explicit
synchronization.
\end{abstract}
\maketitle

\input{sec/introduction}
\input{sec/motivation}
\input{sec/syntax}
\input{sec/types}
\input{sec/semantics}
\input{sec/translation}
\input{sec/algorithmic}
\input{sec/metatheory}
\input{sec/related}
\input{sec/conclusion}

\bibliography{popl2027}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
