\section{Algorithmic Typing}
\label{sec:algorithmic-typing}

\js{Is it ``type variable'' or ``unification variable'' or do we use both
interchangeably?}

This section develops an algorithmic version of the type system in
\cref{sec:types}. The algorithmic system is directly amenable to construct a
type checking algorithm. It employs a sound heuristic for inferring the
remainder session types of split operations. First, a set of constraints is
collected involving fresh type variables. In a second step, the type variables
are solved and validated in respect to the constraints. The constraint
generation is part of the algorithmic typing judgements
(\cref{sec:alg-typing-relation}). How constraint set are solved is discussed in
\cref{sec:constraint-solving}.


\subsection{Algorithmic Judgments}
\label{sec:alg-typing-relation}

Algorithmic typing requires an extended grammar of session types
$\AS$ that includes unification variables $\TVar$ drawn from
$\TVAR$. We write $\AT$ for types that refer to $\AS$ instead of $\S$.

Algorithmic typing collects a set of constraints $\CSet$. Constraints $C$
come as either equality constraints between two types, or mobility constraints:
\[
  C \grmeq* \AT \CEq \ATU \grmor\CMobile \AT
\]

\begin{definition}[Closing substitution]
  A substitution $\sigma : \TVAR \to \AS$ is a \emph{closing
    substitution} for  a set of constraints $\CSet$, written $\sigma
  \models \CSet$, if for every variable $\TVar$ in $\CSet$, $\sigma (\TVar) \in \S$ and all
  constraints are respected.
\end{definition}
\begin{figure}[tp]
  \begin{minipage}{0.49\linewidth}
  \[ \Ctx[1] \CtxJoin \Ctx[2] = \begin{cases}
        \Ctx[1] \CtxPar \Ctx[2] & \text{if $\Dir \in \{ \DirUnord, \DirNone \}$} \\
        \Ctx[1] \CtxSeq \Ctx[2] & \text{if $\Dir = \DirLeft$} \\
        \Ctx[2] \CtxSeq \Ctx[1] & \text{if $\Dir = \DirRight$}
     \end{cases}
  \]
  \end{minipage}
  \begin{minipage}{0.49\linewidth}
\[
  \CtxRestrict*{X} = \begin{cases}
    \CtxRestrict*[1]{X} \CtxPar \CtxRestrict*[2]{X} & \text{if $\Ctx=\Ctx[1] \CtxPar \Ctx[2]$} \\
    \CtxRestrict*[1]{X} \CtxSeq \CtxRestrict*[2]{X} & \text{if $\Ctx=\Ctx[1] \CtxSeq \Ctx[2]$} \\
    \CtxVar \EVar \T & \text{if $\Ctx=\CtxVar \EVar T$ and $\EVar \in X$} \\
    \CtxEmpty & \text{otherwise}
  \end{cases}
\]
  \end{minipage}
  \caption{Context join $\Ctx[1] \CtxJoin \Ctx[2]$ joins two contexts according to the
    direction $\Dir$ and context
    restriction $\CtxRestrict*{X}$ restricts a context to a set of variables $X$}
  \label{fig:context-operators}
\end{figure}
\begin{figure}
  \small
  \hfill \fbox{$\InferType \E \T \Eff$}
  \begin{mathpar}
    \RuleAlgVar \and
    \RuleAlgConst \and
    \RuleAlgAnnot \and
    \RuleAlgLSplit \and
    \RuleAlgRSplit \and
    \RuleAlgSelect \and
    \RuleAlgBranch \and
    \RuleAlgApp \and
    \RuleAlgSeq \and
    \RuleAlgLetPair \and
    \RuleAlgCase
  \end{mathpar}

  \vspace{1em}
  \hfill \fbox{$\CheckType \E \T \Eff$}
  \begin{mathpar}
    \RuleAlgAbs \and
    \RuleAlgAbsRec \and
    \RuleAlgInj \and
    \RuleAlgPair \and
    \RuleAlgCheck
  \end{mathpar}
  \caption{Algorithmic typing}
  \label{fig:typing-algorithmic}
\end{figure}

\Cref{fig:typing-algorithmic} presents the algorithmic typing
judgments, which are based on bidirectional typing~\cite{DBLP:journals/csur/DunfieldK21}. The
synthesis judgment $\InferType \E \AT \Eff$ takes as input the context $\Ctx$
and expression $\E$. It computes the expression's type $\AT$, its effect
$\Eff$, and a set of constraints $\CSet$. The checking judgment $\CheckType \E
\AT \Eff$ takes as input the context $\Ctx$, expression $\E$, and type $\AT$. It
returns the expression's effect $\Eff$ as well as a set of constraints
$\CSet$. Judgments make use of the context operators defined in
\cref{fig:context-operators}.

We have checking rules for function definitions, pairs, sum
injections, and checking against a synthesized type. For lambdas,
\RuleName{A-Abs} enforces the supplied function type, which specifies
the mobility $\Mob$, the direction $\Dir$, and the effect $\Eff$. If
$\Mob = \MobMobile$, then all free variables in the lambda's body must
be mobile. The direction enforces the way the new binding is joined
to the context. The effect puts an upper bound on the effect of the
body. The mobility constraints are inserted into the outgoing
constraint set $\CSet$. \RuleName{A-AbsRec} is similar, but specialized to a recursive
function, which is unrestricted and hence mobile.
For pairs, \RuleName{A-Pair} enforces the ordering of the pair; for
injections, \RuleName{A-Inj} specifies the types of the remaining
summands. Rule \RuleName{A-Check} switches between checking and
synthesis as usual, but leaves checking the equality of the inferred
type $\TU$ against the checked type $\T$ as a constraint.
The other rules (and most synthesis rules) just gather constraints
from the subterms and propagate them.

In the synthesis rules, the context in the variable rule
\RuleName{A-Var} must subsume to a single variable binding and it must be
empty for most constants. A type annotation (\RuleName{A-Annot})
$\E:\T$ synthesizes $\T$ and checks $\E$ against $\T$. Rule
\RuleName{A-LSplit} asks for $\CSplit$ to be annotated with the
borrowed type ${\S}$. This annotation defines the output type
$\TArr{\MobMobile}{\DirUnord,\EffPure} {\SSeq\S\TVar}
{\TPairOrd{\S}{\TVar}}$ up to the continuation, which is supplied by a
fresh variable $\TVar$. This rule does \emph{not} add a
constraint. The constraint appears due to the interplay of rules
\RuleName{A-App} and \RuleName{A-Check}: When $\CSplit_{\S}$ is
applied to an argument, the type of the argument is checked against
$\SSeq\S\TVar$. Solving the equation generated by this check resolves
the assignment to $\TVar$.

Rule \RuleName{A-RSplit} is similar. The rules \RuleName{A-Select}
and \RuleName{A-Branch} normalise \cite{DBLP:journals/iandc/AlmeidaMTV22} the type of the argument before checking it is the appropriate choice type.
We extend type normalisation with one additional rule to push sequiential composition into the choice types \Cref{fig:typing-normalisation}.
Function application
\RuleName{A-App} has to cater to two
different styles of function application, which differ in the
evaluation order of their subterms given by some $\Dir \in \{\DirLeft,
\DirRight\}$. The inferred function type's direction $\Dir'$ is
compatible with $\Dir$ if $\reqDir (\Dir', \Dir)$ holds: All
functions, except $\DirRight$ functions are compatible with
left-to-right evaluation. The function's direction $\Dir'$ also
determines which subterm can have effects as prescribed by $\reqEff
(\Dir', \Eff[1], \Eff[2])$.
\begin{align*}
  \reqDir & = 
  \{(\DirLeft,\DirLeft),
    (\DirRight,\DirRight),
    (\DirUnord,\DirRight),
    (\DirNone,\DirRight)\}
  &
  \reqEff & =
  \{(\DirLeft,\EffPure,\_),
    (\DirRight,\_,\EffPure)\}
\end{align*}
Apart from that, rule \RuleName{A-App} synthesizes the function type and
check against the argument type, as standard. Sequential execution
requires that the sequence of the required contexts for the
expressions can be subsumed from the incoming context. Rule
\RuleName{A-LetPair} considers its subexpressions in sequence and
includes the variable binding in the context of the body according to
the order $\Dir$ of the pair.
Elimination of a sum type using \RuleName{A-Case} works similarly, but
in addition we have to add equations to $\CSet$ that force all branch
output types to be equal.

\subsection{Constraint Solving}
\label{sec:constraint-solving}

An algorithmic typing for an expression $\E$ yields a set of constraints
$\CSet$. In inference mode, the returned type $\AT$ may refer to unsolved type
variables. Solving $\CSet$ and determining valid instantiations for all type
variables boils down to a unification problem. Unfortunately, there is no
unification algorithm readily available for the theory of context-free session
types, which involves associativity, neutral elements, and distributivity. For
this reason, we approach constraint solving by constraint simplification.

The equivalence constraints in $\CSet$ are broken down according to the
simplification rules given in \cref{fig:constraint-simplification} until no
more rules apply. Constraints of the form $\TVar\CEq\AT$ provide a potential
solution for type variable $\TVar$. If there are type variables occurring in
$\CSet$ or $\AT$ for which no solution candidate can be discovered in this way
the heuristic will fail and $\E$ is considered ill-typed.

\begin{figure}[tp]
  \begin{mathpar}
    \RuleCSPair \and
    \RuleCSSum \and
    \RuleCSUnit \and
    \RuleCSArrow \and
    \RuleCSContext
  \end{mathpar}
  \caption{Constraint simplification}
  \label{fig:constraint-simplification}
\end{figure}

The potential solutions have to be validated in respect to the whole set of
constraints $\CSet$. One candidate is picked per variable and substituted
giving a variable-free $\CSet*$\!. The mobility constraints are easily
validated since for any type $\T$ it is decidable if $\Mobile\T$. The
equalities are passed to FreeST which implements a decision procedure for
equivalence of CFSTs. Since the type language of this paper contains constructs
not part of FreeST (e.\,g. the distinction between left and right-ordered
function types) an encoding from our types to FreeST is necessary.
\js{reference to the encoding} If there are constraints that are not satisfied
at this point the program is considered ill-typed. \js{refering some $\E$ above
and ``program'' here, need to fix this}

If there are multiple candidates for a type variable the type to be substituted
into $\CSet$ can be picked non-deterministically. Consider the case where after
simplification there two constraints $\TVar \CEq \T$ and $\TVar \CEq \TU$. If
$\T$ is not a valid solution for $\TVar$, then $\TU$ will not be one either.
The equality $\T \simeq \TU$ must hold whichever type is picked. Since mobility
respects type-equivalence it is not possible to construct mobility constraints
that are satisfied by one but not the other candidate.

In the following we sketch why our heuristic is able to infer useful
information. Fresh type variables are only introduced through
\RuleName{A-LSplit} and \RuleName{A-RSplit}. In the case for $\CSplit$ the
fresh variable $\TVar$ occurs nested in the function result type not composed
with other session types. The function itself is linear, guaranteeing that a
value of the result type will be lying around. \js{continue here}

$\SSkip$ as well as any other skipping session type \js{do we use this
terminology?} are never valid solutions for unification variables as this would
be equivalent to splitting a session type from itself.

\subsection{Implementation}

We implemented a type checker for \thecalculus in Rust, via extending the implementation from \cite{saffrich25:_artif_borrow_from_session_types}. The implementation supports all the features from the paper, except that $\CSend$ and $\CRecv$ require type annotations on the sent/received value. This is because when a channel's type is a unification variable, the return type of $\CRecv$ can't be inferred, so the user must annotate the channel's type. \os{Is it too much detail?} 

We use FreeST \cite{DBLP:journals/corr/abs-1904-01284,DBLP:journals/iandc/AlmeidaMTV22, almeida26:_frees_concur_progr_languag_with} 5.0 to perform equivalence checking in the implementation. \os{We already mention that we use FreeST in Equivalence checking section. Do we need to mention it here again?}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../popl2027.tex"
%%% End:
