\section{Metatheory}
\label{sec:metatheory}

This section summarizes the main metatheoretic guarantees of \thecalculus.
We first state type safety for declarative typing, then weak operational
correspondence for the translation, and finally soundness and completeness of
algorithmic typing.

% \subsection{Structural Properties}
% \begin{lemma}[Unrestricted decomposition]
%   Every context $\Ctx$ can be separated into contexts $\Ctx[1], \Ctx[2]$ with
%   $\Ctx = \Ctx[1] \CtxPar \Ctx[2]$ such that $\Unr\Ctx[1]$ and $\Unr\T$ is
%   not derivable for any $\CtxVar\EVar\T \in \Ctx[2]$.
% \end{lemma}
% \begin{proof}
%   \texttt{unjoinUnr}
% \end{proof}
% 
% \begin{lemma}[Monoidal properties of contexts]
%   The context constructors satisfy the following properties with respect to
%   context equality.
%   \begin{itemize}
%     \item Sequential composition \textup{($\Ctx[1] \CtxSeq \Ctx[2]$)} forms a
%       monoid with identity $\CtxEmpty$.
%     \item Parallel composition \textup{($\Ctx[1] \CtxPar \Ctx[2]$)} forms a
%       commutative monoid with identity $\CtxEmpty$.
%   \end{itemize}
% \end{lemma}
% 
% \begin{lemma}[Unrestricted contexts are mobile]
%   $\Unr\Ctx \implies \Mobile\Ctx$.
% \end{lemma}
% \begin{proof}
%   By induction over $\Ctx$ and the types of bound variables therein.
% \end{proof}

\subsection{Type Safety for Declarative Typing}

For expressions, progress includes a concurrency-specific blocked case. Informally,
an expression is blocked when evaluation reaches a communication primitive in an
evaluation context that cannot proceed on its own because it requires a matching
action from another thread. The formal blocked-expression predicate is given in
the supplementary material in \cref{fig:blocked-expr-proc}. We write $\SessCtx$
for a context $\Ctx$ that only contains session types.

\begin{theorem}[Expression progress\;\AgdaBird]
  If $\HasType \SessCtx \E \T$, then either $\E$ is a value or $\E$ is blocked on concurrent actions or there exists an $\E*$ such that $\E \EReduce \E*$.
\end{theorem}

A process is concurrently blocked, if the active threads are blocked and no
channel pair can synchronize. The formal blocked-process
predicate also is given in the supplementary material in \cref{fig:blocked-expr-proc}.
Process progress guarantees reduction only for non-terminal, non-blocked
configurations. Thus, deadlocks remain possible.

\begin{theorem}[Process progress\;\AgdaBird]
  If $\ProcHasType \SessCtx \Proc$, then either $\Proc \Congruent \PExp{\CUnit}$ or an expression in $\Proc$ is blocked on concurrent actions or
    there exists an $\Proc*$ such that $\Proc \PReduce \Proc*$.
\end{theorem}

Progress for expressions and processes is complemented by preservation,
yielding the usual safety invariant across reduction steps.

\begin{theorem}[Expression preservation\;\AgdaBird]
  If $\HasType \SessCtx \E \T$ and $\E \EReduce \E*$, then
  $\HasType \SessCtx {\E*} \T$.
\end{theorem}

\begin{theorem}[Process preservation\;\AgdaBird]
  If $\ProcHasType \SessCtx \Proc$ and $\Proc \PReduce \Proc*$, then
  $\ProcHasType \SessCtx {\Proc*}$.
\end{theorem}

\subsection{Weak Operational Correspondence for Translation}

We can relate source reduction and target reduction. Since the translation
introduces administrative synchronization steps, correspondence is formulated
as a weak operational correspondence up to zero or more target reductions and congruence.

\begin{theorem}[Forward simulation\;\AgdaBird]
  Let $\Sub$ be a value substitution.
  Additionally, let $\Proc \TransProc[\Sub] \TProc*$ and $\ProcQ \TransProc[\Sub] \TProcQ*$.
  If $\ProcHasType \SessCtx \Proc$ and $\Proc \PReduce \ProcQ$, then
  either $\TProc* \TPReduce^{*} \TProcQ*$ or $\TProc* \Congruent \TProcQ*$.
\end{theorem}

The converse direction states that every target step can be matched by a finite
source computation, again modulo weak closure and congruence.

\begin{theorem}[Backward simulation\;\AgdaBird]
  Let $\Sub$ be a value substitution.
  Additionally, let $\Proc \TransProc[\Sub] \TProc*$.
  If $\ProcHasType \SessCtx \Proc$ and $\TProc* \TPReduce \TProcR*$, then there
  exist $\ProcQ$ and $\TProcQ*$ such that $\Proc \PReduce^{*} \ProcQ$ and
  $\TProcR* \TPReduce^{*} \TProcQ*$ where $\ProcQ \TransProc[\Sub] \TProcQ*$.
\end{theorem}
\subsection{Algorithmic Typing}

We now connect algorithmic typing with declarative typing. Soundness says
that any solved algorithmic judgment denotes a valid declarative typing, while
completeness says that declarative typings can be reconstructed algorithmically.

\begin{theorem}[Expression checking soundness\;\AgdaBird]
  If $\CheckType \E \AT \Eff$ and $\sigma \models \CSet$ is a closing
  substitution such that $\Fv(\AT) \subseteq \dom(\sigma)$, then
  $\HasType \Ctx \E {\AT{[\sigma]}}$.
\end{theorem}

The same soundness statement holds for inference mode.

\begin{theorem}[Expression inference soundness\;\AgdaBird]
  If $\InferType \E \AT \Eff$ and $\sigma \models \CSet$ is a closing
  substitution such that $\Fv(\AT) \subseteq \dom(\sigma)$, then
  $\HasType \Ctx \E {\AT{[\sigma]}}$.
\end{theorem}

For completeness, we again first cover checking and then synthesis.

\begin{theorem}[Expression checking completeness\;\AgdaBird]
  If $\HasType \Ctx \E \T$, then
  $\CheckType \E \AT \Eff$ for some type $\AT$, set of constraints $\CSet$,
  and closing substitution $\sigma$ such that $\T = \AT{[\sigma]}$.
\end{theorem}

\begin{theorem}[Expression inference completeness\;\AgdaBird]
  If $\HasType \Ctx \E \T$ and $\E$ is in synthesis form, then
  $\InferType \E \AT \Eff$ for some type $\AT$, set of constraints $\CSet$,
  and closing substitution $\sigma$ such that $\T = \AT{[\sigma]}$.
\end{theorem}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../popl2027.tex"
%%% End:
