
\section{Motivation}
\label{sec:motivation}

Context-free session types (CFSTs) extend regular session types with a
general sequential composition operator. A type $S_1;S_2$ describes a
session that first behaves as $S_1$ and then continues as $S_2$. The
inactive session \lstinline{Skip} is the unit of this composition and
composition is associative.
% , that is, \lstinline{Skip};$S = S =
% S$;\lstinline{Skip} and $(S_1;S_2);S_3 = S_1;(S_2;S_3)$.
Thus
\lstinline{Skip} and \lstinline{;} form a monoid up to type
equality. The types \lstinline{Close} and \lstinline{Wait} indicate
(active and passible) termination of a session.

Sequential composition also interacts with choice in the expected way.
Appending a continuation to a choice appends it to every branch:
$(\oplus\{\ell_i:S_i\}_{i\in I});T =
\oplus\{\ell_i:S_i;T\}_{i\in I}$. The same law holds for external
choice. These equations are what make non-tail-recursive protocols
usable as ordinary session types: a continuation can be pushed through
the structure of a type until it reaches the leaves of a protocol.

As usual for session types, channels are linear resources. A channel
has to be used exactly once and every communication operation consumes
the channel at its current type and produces the channel at its
continuation type. Linearity is what makes the session type track the
state of the protocol, but it also forces programs to thread channels
explicitly from one operation to the next.

\subsection{Borrowing from Context-Free Session Types}
\label{sec:borr-from-cont}


Context-free session types enable writing elegant code for
serializing tree structures when communication is restricted to
primitive values.
% \footnote{We assume session-typed communication over a network
%   or between sandboxed processes, where only primitive data can be
%   passed, but no pointers.}
Consider an example from \citet{DBLP:conf/icfp/ThiemannV16} for serializing binary trees of type \lstinline{Tree}.
\begin{lstlisting}
type Tree = Leaf
          | Node int Tree Tree
\end{lstlisting}
Here is the corresponding session type \lstinline{TreeChannel}:
\begin{lstlisting}
type TreeChannel = oplus{
  Leaf: Skip,
  Node: !int; TreeChannel; TreeChannel
}
\end{lstlisting}
\begin{figure}[tp]
  \begin{subfigure}[t]{0.49\linewidth}
\begin{lstlisting}
send : forallalphasigma. alpha -> !alpha;sigma -> sigma
recv : forallalphasigma. ?alpha;sigma -> alphaoxsigma
select ellj : +{elli: Si} -> Sj
branch : &{elli: Si} -> [elli: Si]
\end{lstlisting}
    \caption{CFST API}\label{lst:cfst-api}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\linewidth}
\begin{lstlisting}[style=eclipse-new]
send : forallalpha. alpha -> !alpha -> Unit
recv : forallalpha. ?alpha -> alpha
select ellj : +{elli: Si} -> Sj
branch : &{elli: Si} -> [elli: Si]
\end{lstlisting}
    \caption{\thecalculus API}\label{lst:bcfst-api}
  \end{subfigure}
  \caption{Session type APIs}
  \label{fig:session-type-apis}
\end{figure}
The type \lstinline{TreeChannel} is called context-free because it is
recursively defined and the recursion is not restricted to tail
positions. The \lstinline{Node} alternative is an example where the
\lstinline{TreeChannel} type refers two times in a row to itself. The
first occurrence of \lstinline{TreeChannel} in this
type is illegal in regular session type systems (e.g.,
\cite{DBLP:journals/jfp/GayV10,DBLP:conf/esop/LindleyM15,Gay_Vasconcelos_2025}) because
they syntactically restrict recursion to tail position. Context-free
session types lift this restriction as their composition operator $S_1;S_2$ connects
two sessions sequentially: the composed session comprises the
actions of $S_1$ followed by the actions of $S_2$.
% \footnote{One
%   might argue that CFSTs are close to Honda's original vision for
%   session types \cite{DBLP:conf/concur/Honda93}, which also features
%   symmetric composition.}

\Cref{lst:cfst-api} displays the API of CFSTs. The \lstinline{send},
\lstinline{recv}, and \lstinline{select} operations are standard for
sending a message, receiving one, and selecting an alternative in a
branching session type. The \lstinline{branch} operation maps a channel
offering different labeled alternatives to a variant type with the
same labeled alternatives. Standard pattern matching suffices to
decode the chosen alternative and continue.
Given the types and the CFST API, we can implement the serialization function:
\begin{lstlisting}[numbers=left]
sendTree : Tree -> forallsigma.TreeChannel;sigma -> sigma
sendTree Leaf c =
  select Leaf c
sendTree (Node x l r) c =
  let c1 = select Node c   (* c1 : !int; TreeChannel; TreeChannel; sigma *)
      c2 = send x c1       (* c2 : TreeChannel; TreeChannel; sigma *)
      c3 = sendTree l c2   (* c3 : TreeChannel; sigma *)
      c4 = sendTree r c3   (* c4 : sigma *)
  in  c4
\end{lstlisting}
However, there are several catches. First, the channel \lstinline{c}
is passed explicitly throughout the code; a pattern we call
\emph{resource-passing style}. This style forces the programmer to
make the function polymorphic on the continuation type of the channel,
which is \lstinline{sigma} in this example.
Second, the \lstinline{Node} branch of \lstinline{sendTree} relies on
polymorphic recursion. To see this, consider the type of
\lstinline{c2} in line 6--7. To invoke \lstinline{sendTree}
recursively, we need to instantiate the type argument
\lstinline{sigma} with \lstinline{TreeChannel; sigma} rather than just
\lstinline{sigma}, which is the
telltale for polymorphic recursion.

Our new language {\thecalculus} amends these problems. The type
definitions remain unchanged, but the implementation of
\lstinline{sendTreeB} relies on the session API with borrowing shown in
\cref{lst:bcfst-api}: the \lstinline{send} and \lstinline{recv} adopt
the borrowing style, but \lstinline{select} and \lstinline{branch}
remain in resource-passing style (see \cref{sec:remarks} for the
reason). To make this switch apparent, we present code
using the \thecalculus API with grey background and golden keywords.
\lstset{style=eclipse-new}
\begin{lstlisting}[numbers=left]
sendTreeB : Tree -> TreeChannel -> Skip
sendTreeB Leaf c =          (* c : TreeChannel *)
  select Leaf c
sendTreeB (Node x l r) c =
  let c = select Node c in  (* c : !Int; TreeChannel; TreeChannel *)
  send x &c;                (* c : TreeChannel; TreeChannel *)
  sendTreeB l &c;           (* c : TreeChannel *)
  sendTreeB r &c
\end{lstlisting}
What has changed? First, the type of \lstinline{sendTreeB} is no longer polymorphic over the
continuation type. Instead, the channel argument \lstinline{c} starts
out at type \lstinline{TreeChannel} without specifying a
continuation. The return type is \lstinline{Skip}, which indicates an inactive
channel. Values of type \lstinline{Skip} are
unrestricted and can be discarded.
Second, there is no explicit resource
passing, except for \lstinline{select}. Each use of the borrow notation
\lstinline{&c} implicitly splits off a prefix of the session type
according to the function argument and updates \lstinline{c} to the
suffix. To see this, consider the type of \lstinline{c}---after
splitting off the borrow in the corresponding line---shown in the
comments. Before the \lstinline{send} operation (line 5--6), the type
of \lstinline{c} starts with \lstinline{!Int; ...}. ``Taking the
borrow'' splits off the borrowed type \lstinline{!Int}---the
argument type of \lstinline{send x}---and
updates \lstinline{c} to the type shown in line 6. As the type of
\lstinline{c} is itself a borrowed type, one could speak of a
\emph{reborrow}. 

Third, as there is no polymorphism, there is no polymorphic
recursion, either.

Finally, the channel variable \lstinline{c} can be used in a
non-linear way, as long as its uses are under borrows, as in
\lstinline{&c}. The understanding is that the borrows are taken and
processed according to the evaluation order of the program. This view does
not abandon linearity. Rather, the type system internalizes the
linear use of the channel: each borrow consumes a distinct prefix of
the current session type and leaves the residual type for the next use
of \lstinline{c}.


\subsection{Local and Remote Borrowing}
\label{sec:local-remote-borr}

\Cref{sec:borr-from-cont} demonstrated
 \emph{local borrowing}, where the borrows remain in the same
thread and they are processed sequentially, following the evaluation order.

In this local setting, operations on borrows are
cheap: borrowing merely names the next prefix of the same channel and
leaves the suffix for the surrounding computation. No other process can
race with the continuation, so the evaluation order of the program and
the left-to-right order enforced by typing already provide the required
sequencing.

However, in some situations we want to send a borrow to
another process. As an example, suppose a server-side renderer streams
HTML nodes to a client instead of constructing a DOM tree in
memory first. A small protocol for such a renderer is as follows.
\begin{lstlisting}
type HtmlChannel = oplus{
  Text: !String,
  Elem: !String; ChildrenChannel
}
type ChildrenChannel = oplus{
  Done: Skip,
  Child: HtmlChannel; ChildrenChannel
}
\end{lstlisting}
The \lstinline{Elem} alternative first sends a tag name, and then a sequence of children. The 
\lstinline{Child} alternative is context-free in the same sense as the
\lstinline{Node} alternative of \lstinline{TreeChannel}: rendering one
child consumes an \lstinline{HtmlChannel} prefix and then continues with
the rest of the child stream.

Now consider a renderer (\Cref{lst:rendering-example}) that emits a user panel whose first child is
computed by another process while the main thread obtains user
messages, say, from a database. We assume that
\lstinline{renderProfile} and \lstinline{renderMsgArea} render
one HTML node on a channel of type \lstinline{HtmlChannel; Ret} and
\lstinline{HtmlChannel; Close}, respectively. The type \lstinline{Ret}
is a new session type that comes with a companion type
\lstinline{Acq}. They indicate the placement of synchronization
operations that ensure correct handover of ownership from the borrowed session
to the session's suffix.
\begin{lstlisting}[numbers=left,float,caption={Rendering example},captionpos=b,label={lst:rendering-example}]
renderUser : User -> HtmlChannel; Close -> Unit
renderUser user c =
  let c = select Elem c in
  send "div" &c;
  let c = select Child c in
  fork (lambda() -> renderProfile user &c);
  (* c : Acq; ChildrenChannel; Close *)
  let msgs = getMessages user in
  renderRest msgs (acquire c)

renderRest : Messages -> ChildrenChannel; Close -> Unit
renderRest msgs c =
  let c = select Child c in
  renderMsgArea msgs &c;
  close (select Done c)

renderProfile : User -> HtmlChannel;Ret -> Unit
renderProfile user c = ... drop c
\end{lstlisting}
The function \lstinline{renderUser} starts an HTML element and then
opens the child stream. It passes a borrow of the
first child position to a new process, which renders the user's profile.
The continuation renders the remaining children, here a messaging
area, and then closes the child stream. 

In this case, the borrow under the \lstinline{fork} is a \emph{remote
  borrow}. Such a borrow must be handled differently as operations on
the borrow are not necessarily ordered, anymore: the 
continuation must \emph{not} start emitting the second child before the
forked renderer has finished the first child. This synchronization is
expressed with the \lstinline{acquire} operation, which works on a
channel of type \lstinline{Acq; ...}. It blocks the
continuation until the borrowed prefix has been completed by the
forked process. Conversely, the forked process calls \lstinline{drop}
on a channel of type \lstinline{Ret} when it has finished working on
the prefix. Thus \lstinline{drop} and
\lstinline{acquire} form the synchronization point between the two
processes; the run-time mechanism that implements this point is
described in~\cref{sec:impl-local-remote}.

This setup is also reflected in the types that result from a remote
borrow. Line~7 shows the type of \lstinline{c}'s suffix, which starts
with an \lstinline{Acq}, and line~16 shows the type of the borrow,
which ends with \lstinline{Ret}. The types \lstinline{Acq} and
\lstinline{Ret} only arise from a remote split operation. The
\emph{only} operation on a channel of type
\lstinline{Acq; ...} is \lstinline{acquire} and similarly,
\lstinline{drop} only applies to a channel of type \lstinline{Ret}. 

Next, we discuss a direct operational semantics of local and remote borrowing
before we move on to a low level implementation of the two concepts in
\cref{sec:impl-local-remote}.

\subsection{A Direct Semantics for Channel Borrowing}
\label{sec:direct-semant-chann}

\Cref{sec:borr-from-cont} already suggests an implementation strategy
for local borrowing. However, we start with an abstract
specification because it is easier to handle in a correctness
proof. The first step translates the surface 
language with borrowing operators into our formal calculus
\thecalculus, which replaces borrowing with explicit functional splitting
operators. This translation is analogous to a translation in previous work
\cite{DBLP:journals/pacmpl/SaffrichS0V25}, which gives a formal
presentation along with a correctness proof. The novelty in our work
is the presence of \emph{two} splitting operators, one for local borrows and
one for remote borrows. Our discussion starts with the local variant and
we show the translation of \lstinline{renderRest} in
\thecalculus. The translation is part of our implementation.

As an example, we use the local part of
\lstinline{renderRest} shown in
\cref{fig:local-split-program}.
Execution proceeds by repeatedly exposing the next local prefix of the
channel sequence. \Cref{fig:local-split-trace} shows a suffix of its reduction
sequence.
The \lstinline{nu[c][d] in ...} operator is a channel abstraction as known
from session calculi \cite{Gay_Vasconcelos_2025}. It introduces a
communication channel with endpoint names \lstinline{c} and 
\lstinline{d}. We generalize channel abstraction so that each
compartment \lstinline{[c]} and \lstinline{[d]} contains a
\emph{binding group}, which binds one or more names. Its working is
best illustrated with the example trace.

To keep the trace short (\cref{fig:local-split-trace}), the
reduction starts after the initial local splits; the omitted steps
only expose the prefix consumed by 
\lstinline{renderMsgArea}.
\begin{figure}
\begin{subfigure}[t]{0.49\linewidth}
\begin{lstlisting}[numbers=left]
K(c) =
  let c2 = select Child c in
  let c3, c4 = lsplit c2 in
  renderMsgArea user c3;
  close (select Done c4)
\end{lstlisting}
\caption{Translated program.}
\label{fig:local-split-program}
\end{subfigure}\hfill
\begin{subfigure}[t]{0.49\linewidth}
\begin{lstlisting}
nu[c][d..] in
  (P_d ||
   let c2 = select Child c in
   let c3, c4 = lsplit c2 in
   renderMsgArea user c3;
   close (select Done c4)
~> nu[c2][d..] in
  (P_d ||
   let c3, c4 = lsplit c2 in
   renderMsgArea user c3;
   close (select Done c4)
~> nu[c3;c4][d..] in
  (P_d ||
   renderMsgArea user c3;
   close (select Done c4)
~> nu[c4][d..] in
  (P_d ||
   close (select Done c4)
~> nu[c4][d..] in
  (P_d ||
   close c4
~> nu[c4][d3] in
  (P_d'[wait d3] ||
   close c4)
~> (P_d'[*] || *)
\end{lstlisting}
\caption{Reduction suffix.}
\label{fig:local-split-trace}
\end{subfigure}
\caption{Local splitting in the suffix of \lstinline{renderRest}.
The processes $P_d$ and $P_d'$ abbreviate client-side
computation on the other endpoint.}
\label{fig:local-split-reduction}
\end{figure}
The split on \lstinline{c2} following the \lstinline{select} \emph{replaces} the
binder for \lstinline{c2} with two fresh channel names \lstinline{c3}
and \lstinline{c4}. That is, both binding compartments of the
\lstinline{nu} contain a \emph{list-like structure} of binders for
channel names and all operations
work on these channel names from left to right. To ensure the correct
sequencing, the semantics does not allow any action on \lstinline{c4} before
\lstinline{c3} has been consumed. This restriction is supported by
typing: the \lstinline{lsplit} operation returns an \emph{ordered
  pair}, whose components must be processed and consumed in
left-to-right order.
% The \lstinline{drop} operation (not used in this
% example) applied to some \lstinline{c} just discards \lstinline{c} if
% it is the first channel name in the compartment.

On the client side, in the process $P_d$ that holds the other
channel end \lstinline{d}, 
splitting can be performed on \lstinline{d} independently of
the splitting on \lstinline{c}. The only constraint is that, in the
end, when \lstinline{close} and \lstinline{wait} calls match, the
reduction requires that only one channel name is left in each
compartment. Under this condition, the channel can be closed and
eliminated. The figure does not show the transitions of the client $P_d$.

For remote borrowing, we incorporate the 
\lstinline{acquire} primitive. Hence, we indicate a remote split with the
symbol ``\lstinline{|}'' in the compartment. As an example, we consider the translation of
lines 5--7:
\begin{lstlisting}
  fork (lambda() -> renderProfile user &c);
  let msgs = getMessages user in
  renderRest msgs (acquire c)
\end{lstlisting}
It translates to the following \thecalculus code.
\begin{lstlisting}
(* initially                  c : HtmlChannel; ChildrenChannel; Close *)
let c1, c2 = rsplit c in   (* c1 : HtmlChannel; Ret *)
                           (* c2 : Acq; ChildrenChannel; Close *)
fork (lambda() -> renderProfile user c1);
let msgs = getMessages user in
renderRest msgs (acquire c2)
\end{lstlisting}
The translation inserts an \lstinline{rsplit} because \lstinline{c1}
and \lstinline{c2} are used in different processes. As their uses are
explicitly synchronized, the \lstinline{rsplit} returns a normal
(linear) pair of channels. The type of the second channel
\lstinline{c2} starts with an \lstinline{Acq} type, a special form
of session type that indicates the requirement for explicit
synchronization.
Executing this code takes the following steps (ignoring the activity
at the other end of the channel). The bracketed step abbreviates the
forked process running \lstinline{renderProfile} until its final
\lstinline{drop c1}.
%     nu[c][...]
%     let c = select Child c in
%     let c1, c2 = rsplit c in
%     fork (lambda() -> renderProfile user c1);
%     let msgs = ... in
%     renderRest msgs (acquire c2)
% ~>

\begin{lstlisting}
    nu[c][...]
    let c1, c2 = rsplit c in
    fork (lambda() -> renderProfile user c1);
    let msgs = ... in
    renderRest msgs (acquire c2)
~>
    nu[c1|c2][...]
    fork (lambda() -> renderProfile user c1);
    let msgs = ... in
    renderRest msgs (acquire c2)
~>*  [ drop c1 ]
    nu[|c2][...]
    renderRest msgs (acquire c2)
~>
    nu[c2][...]
    renderRest msgs c2
\end{lstlisting}
In summary, the \lstinline{rsplit} inserts a new binding group \lstinline{[c1|c2]}
demarcated by ``\lstinline{|}'' to indicate a remote split. The
\lstinline{drop} is only enabled 
if its argument is the last name in a group directly before a
``\lstinline{|}''. The \lstinline{acquire} is only
enabled if the preceding binding group has been processed as in
\lstinline{[|c2]}. Executing the \lstinline{acquire} in the last step removes the
exhausted binding group and yields \lstinline{[c2]}.

\subsection{Implementing Local and Remote Borrowing}
\label{sec:impl-local-remote}

We implement the high-level calculus by translation into
a low-level run-time session calculus with borrowing, \thertcalculus. The run-time calculus
is a concurrent 
lambda calculus with the usual session-style communication
primitives. However, \thertcalculus is untyped and it comes equipped with
asynchronous signaling primitives for implementing \lstinline{drop}/\lstinline{acquire}.
The direct semantics from~\cref{sec:direct-semant-chann} records ordering constraints in binding
groups. The run-time translation realizes the same constraints with
ordinary channel names and optional synchronization flags that
implement passing of ownership.
% \Cref{fig:borrow-lifecycle} summarizes the difference between local
% and remote borrowing in the implementation.

% \begin{figure}[t]
% \centering
% \begin{tikzpicture}[
%   >=Stealth,
%   node distance=0.8cm and 1.4cm,
%   state/.style={draw, rounded corners, align=center, inner sep=3pt, font=\small},
%   split/.style={draw, rectangle split, rectangle split parts=2,
%     rectangle split part align={center,center}, rounded corners,
%     align=center, inner sep=3pt, font=\small},
%   local split/.style={split, rectangle split part fill={white,black!12}},
%   sync/.style={draw, rounded corners, align=center, inner sep=3pt, font=\small, fill=black!6},
%   scope/.style={draw, dashed, rounded corners, inner xsep=0.28cm,
%     inner ysep=0.22cm},
%   lab/.style={font=\small}
% ]
%   \node[state] (l0) {$c:S_1;S_2$};
%   \node[local split, right=of l0] (l1)
%     {$c_1:S_1$\nodepart{two}$c_2:S_2$};
%   \node[state, right=of l1] (l2) {$c_2:S_2$};
%   \node[state, right=of l2] (l3) {continue\\with $S_2$};

%   \draw[->] (l0) -- node[lab, above] {\texttt{lsplit}} (l1);
%   \draw[->] (l1) -- node[lab, above] {use $c_1$} (l2);
%   \draw[->] (l2) -- node[lab, above] {use $c_2$} (l3);

%   \node[lab, left=0.35cm of l0] {\textbf{local}};

%   \node[state, below=1.65cm of l0] (r0) {$c:S_1;S_2$};
%   \node[state, right=1.4cm of r0, yshift=0.3cm] (r1a)
%     {$c_1:S_1;\texttt{Ret}$};
%   \node[state, below=0.35cm of r1a] (r1b)
%     {$c_2:\texttt{Acq};S_2$};
%   \node[state, right=0.95cm of r1a] (rret) {$c_1:\texttt{Ret}$};
%   \coordinate[right=1.35cm of rret.east] (rdrop);
%   \node[state, right=3.25cm of r1b.east] (r2) {$c_2:S_2$};
%   \node[state, right=of r2] (r3) {continue\\with $S_2$};
%   \coordinate[right=0.55cm of r1b.east] (racq);

%   \coordinate[right=0.45cm of r0.east] (rfork);
%   \draw[-] (r0.east) -- node[lab, above] {\texttt{rsplit}} (rfork);
%   \draw[->] (rfork) to[out=25,in=180] (r1a.west);
%   \draw[->] (rfork) to[out=-25,in=180] (r1b.west);
%   \draw[->] (r1a) -- node[lab, above] {use $c_1$} (rret);
%   \draw[-{Stealth[open]}] (rret.east) --
%     node[lab, above] {\texttt{drop} $c_1$} (rdrop);
%   \draw[->] (r1b.east) --
%     node[lab, below, pos=0.7] {\texttt{acquire} $c_2$} (r2);
%   \draw[->] (r2) -- node[lab, above] {use $c_2$} (r3);
%   \draw[->, dashed] (rdrop) to[out=-40,in=140]
%     (racq);
%   \node[scope, fit=(r1a)(r1b)(rret)(rdrop)(r2)(r3),
%     inner xsep=0.45cm, inner ysep=0.45cm] (rscope) {};
%   \node[lab, anchor=north west, xshift=2pt, yshift=-2pt]
%     at (rscope.north west) {$\varphi z \mapsto \texttt{drop}$};
%   \node[lab, anchor=south, yshift=0.35cm] at ($(rret.east)!0.5!(rdrop)$)
%     {$z \mapsto \texttt{acq}$};
%   \node[lab, anchor=south, yshift=0.08cm] at ($(r1b.east)!0.7!(r2.west)$)
%     {$z \mapsto \texttt{done}$};

%   \node[lab, left=0.35cm of r0] {\textbf{remote}};
% \end{tikzpicture}
% \caption{Lifecycle of local and remote borrows. Local borrowing
% exposes an ordered prefix and suffix of the same channel, with the
% suffix inactive while the prefix is used. Remote borrowing inserts a
% synchronization flag $z$ whose scope encloses both pieces; dropping
% the prefix enables acquisition of the suffix.}
% \label{fig:borrow-lifecycle}
% \end{figure}

\Cref{fig:borrow-lifecycle-msc} summarizes the difference between local
and remote borrowing in the implementation.

\begin{figure}[t]
\centering
\begin{subfigure}[t]{0.225\linewidth}
\centering
\vspace{0pt}
\begin{tikzpicture}[
  >=Stealth,
  x=0.775cm,
  y=0.48cm,
  head/.style={draw, fill=white, minimum width=1.15cm,
    minimum height=0.42cm, inner sep=1pt, font=\scriptsize},
  state/.style={draw, fill=blue!5, align=center, inner sep=2pt,
    font=\scriptsize, text width=1.45cm},
  inactive/.style={state, fill=black!12},
  lifeline/.style={dashed, black!55},
  activation/.style={draw, fill=black!8},
  msg/.style={->},
  lab/.style={font=\scriptsize}
]
  \draw[black!70] (-1.0,0.55) rectangle (3.0,-8.4);

  \node[head] (lp) at (0,-0.65) {$P$};
  \draw[lifeline] (lp.south) -- (0,-8.15);
  \draw[activation] (-0.07,-1.15) rectangle (0.07,-8.0);

  \node[state] at (1.25,-1.25) {$c:S_1;S_2$};
  \draw[msg] (0.07,-2.25) -- (0.9,-2.25) --
    node[lab, right] {} (0.9,-2.65) --
    node[lab, above, yshift=1ex, xshift=0.25em] {\texttt{lsplit}}
    (0.07,-2.65);
  \node[state] at (1.25,-3.45) {$c_1:S_1$};
  \node[inactive] at (1.25,-4.25) {$c_2:S_2$\\inactive};
  \draw[msg] (0.07,-5.45) -- (0.9,-5.45) --
    node[lab, right] {} (0.9,-5.85) --
    node[lab, above, yshift=1ex] {use $c_1$}
    (0.07,-5.85);
  \node[state] at (1.25,-6.45) {$c_2:S_2$};
  \draw[msg] (0.07,-7.45) -- (0.9,-7.45) --
    node[lab, right] {} (0.9,-7.85) --
    node[lab, above, yshift=1ex] {use $c_2$}
    (0.07,-7.85);
\end{tikzpicture}
\caption{Local borrow}
\label{fig:borrow-lifecycle-msc-local}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.755\linewidth}
\centering
\vspace{0pt}
\begin{tikzpicture}[
  >=Stealth,
  x=0.775cm,
  y=0.52cm,
  head/.style={draw, fill=white, minimum width=1.25cm,
    minimum height=0.45cm, inner sep=1pt, font=\scriptsize},
  flaghead/.style={head, fill=orange!14, draw=orange!65!black},
  state/.style={draw, fill=blue!5, align=center, inner sep=2pt,
    font=\scriptsize, text width=1.8cm},
  flagstate/.style={draw=orange!65!black, fill=orange!14,
    align=center, inner sep=2pt, font=\scriptsize, text width=1.45cm},
  lifeline/.style={dashed, black!55},
  flagline/.style={dashed, orange!65!black},
  activation/.style={draw, fill=black!8},
  flagact/.style={draw=orange!65!black, fill=orange!10},
  msg/.style={->},
  create/.style={->, dashed},
  return/.style={->, dashed},
  lab/.style={font=\scriptsize}
]
  \draw[black!70] (-1.0,0.55) rectangle (12.55,-11.55);

  \node[head] (rp) at (0,-0.65) {$P$};
  \draw[lifeline] (rp.south) -- (0,-11.2);
  \draw[activation] (-0.08,-1.15) rectangle (0.08,-5.15);
  \draw[activation] (-0.08,-8.8) rectangle (0.08,-11.0);
  \node[state] at (1.45,-1.25) {$c:S_1;S_2$};

  \node[flaghead] (rz) at (4.8,-2.35) {$\varphi\,z$};
  \draw[create] (0.08,-2.35) --
    node[lab, above] {\texttt{rsplit}: create $z$}
    (rz.west);
  \draw[flagline] (rz.south) -- (4.8,-11.2);
  \draw[flagact] (4.72,-2.85) rectangle (4.88,-8.8);
  \node[flagstate] at (5.85,-3.15) {$z \mapsto \texttt{drop}$};

  \node[state] at (1.45,-3.55) {$c_2:\texttt{Acq};S_2$};
  \node[head] (rq) at (9.15,-4.25) {$Q$};
  \draw[create] (0.08,-4.25) -- node[lab, above] {fork $c_1$} (rq.west);
  \draw[lifeline] (rq.south) -- (9.15,-8.25);
  \draw[activation] (9.07,-4.7) rectangle (9.23,-7.9);

  \draw[msg] (0.08,-5.15) -- node[lab, above] {\texttt{acquire} $c_2$}
    (4.72,-5.15);
  \node[state, text width=2.2cm] at (10.85,-5.15)
    {$c_1:S_1;\texttt{Ret}$};
  \draw[msg] (9.23,-6.45) -- (10.05,-6.45) --
    node[lab, right] {} (10.05,-6.85) --
    node[lab, above, yshift=1ex] {use $c_1$}
    (9.23,-6.85);
  \draw[msg] (9.07,-7.55) -- node[lab, above] {\texttt{drop} $c_1$}
    (4.88,-7.55);
  \node[flagstate] at (5.85,-8.15) {$z \mapsto \texttt{acq}$};
  \node[flagstate] at (5.85,-8.8) {$z \mapsto \texttt{done}$};
  \draw[return] (4.72,-8.8) -- node[lab, above] {return} (0.08,-8.8);

  \node[state] at (1.45,-9.35) {$c_2:S_2$};
  \draw[msg] (0.08,-10.25) -- (0.9,-10.25) --
    node[lab, right] {} (0.9,-10.65) --
    node[lab, above, yshift=1ex] {use $c_2$}
    (0.08,-10.65);
\end{tikzpicture}
\caption{Remote borrow}
\label{fig:borrow-lifecycle-msc-remote}
\end{subfigure}
\caption{Message-sequence view of local and remote borrows. Channel state is shown as
annotations next to the relevant process or flag lifelines. A local
split keeps the channel state with the original process lifeline; a
remote split creates a flag lifeline $\varphi\,z$ and forks a process
for the borrowed prefix.}
\label{fig:borrow-lifecycle-msc}
\end{figure}

Specifically, a channel is represented by a triple
$in \otimes c \otimes out$, where $c$ is a primitive channel name and
$in$ and $out$ are optional names of synchronization flags. The
incoming flag $in$ records whether the channel must be acquired before
it can be used. The outgoing flag $out$ records whether the channel
must be dropped after use. A synchronization binder
$(\varphi z \mapsto \phi)P$ binds a flag name $z$ in $P$ and stores
one of three states in $\phi$: \lstinline{drop}, \lstinline{acq}, or
\lstinline{done}. An \lstinline{rsplit} creates a flag in state
\lstinline{drop} and a channel creation sets its flag to \lstinline{acq}.
Intuitively, \lstinline{drop} means that the left, borrowed
part of a remote split is still active, \lstinline{acq} means that it
has been dropped and the right, suffix part may acquire, and \lstinline{done}
means that the synchronization has served its purpose and can be
removed. This mechanism implements an ownership transfer of the
primitive channel from the borrowed part to the suffix.

Allocation creates a primitive channel pair and one synchronization
flag for each endpoint. The fresh endpoints carry incoming flags that
are already in the \lstinline{acq} state. Thus, an initial
\lstinline{acquire} is required to reduce each channel to normal
operation, after which a cleanup rule removes the flag from the 
channel triple by replacing it with \lstinline{*}. Once these optional
flag components are absent, communication is just ordinary
session communication: \lstinline{send}/\lstinline{recv},
\lstinline{select}/\lstinline{branch}, and \lstinline{close}/\lstinline{wait}
all operate directly on the primitive channel name $c$.

Local splitting is deliberately cheap (\cref{fig:borrow-lifecycle-msc-local}). The rule for
\lstinline{lsplit} returns two copies of the same run-time channel
triple. No new channel is allocated and no synchronization flag
is introduced. This is sound because the source type system has
already established the left-to-right use of the two pieces. The
run-time representation can therefore be imperative: the two pieces are
aliases for the same underlying channel object, and using the first
piece advances that object to the state where the second piece starts.

Remote splitting is the only place where new synchronization
machinery is introduced
(\cref{fig:borrow-lifecycle-msc-remote}). Splitting a channel
$in \otimes c \otimes out$ remotely creates a fresh flag $z$ in state
\lstinline{drop} and returns two channel triples:
$in \otimes c \otimes z$ for the borrowed prefix, and
$z \otimes c \otimes out$ for the suffix. The prefix therefore carries
an outgoing obligation to call \lstinline{drop}; the suffix carries a
matching incoming obligation to call \lstinline{acquire}. Calling
\lstinline{drop} changes the flag from \lstinline{drop} to
\lstinline{acq} and returns immediately, so the notification is
asynchronous. A later \lstinline{acquire} on the suffix can fire only
after this state change. It changes the flag to \lstinline{done} and
returns the channel. A cleanup rule then removes the synchronization
flag altogether. Thus a remote split pays for exactly the
synchronization needed to coordinate the two processes operating on
the channel, whereas a local
split pays nothing beyond sharing the channel representation.

\subsection{Remarks}
\label{sec:remarks}

The reader may ask why \lstinline{select} still adheres to
resource passing style. The answer is simplicity of our system. While it
is possible to define \lstinline{select} with a borrowing-based
interface, supporting it requires subtyping of session types, which is
undecidable for CFST~\cite{DBLP:journals/toplas/Padovani19} and hence
not implementable.

For borrowing from regular session
types~\cite{DBLP:journals/pacmpl/SaffrichS0V25}, a borrowing-style 
\lstinline{select} operation is possible because subtyping for regular
session types is decidable.

A borrowing-style \lstinline{branch} operator cannot be supported in \thecalculus
because this operator externally selects between different continuations.

% The downside is that some uses of \lstinline{select} force us to use
% the \lstinline{drop} primitive, which discharges an exhausted channel
% borrow of type \lstinline{Ret}. The discharge is part of the
% \lstinline{send} and \lstinline{recv} operations, so it is rarely
% needed in practice. Alternatively, the type checker could insert
% \lstinline{drop} automatically.


All \thecalculus examples are available as part of our supplement. Compared to the
presentation in this section, the split operators carry type
parameters in the actual code to simplify type checking.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../popl2027.tex"
%%% End:
