
\section{Related Work}
\label{sec:related-work}

\subsection{Session Types}
\label{sec:session-types}

Context-free session types (CFSTs) have been introduced by
\citet{DBLP:conf/icfp/ThiemannV16} and further developed by
\citet{DBLP:journals/toplas/Padovani19} and
\citet{DBLP:journals/iandc/AlmeidaMTV22}. CFSTs liberate traditional 
session types from their tail-bias by replacing the prefix type
operators for the session operations by symmetric composition. This
move complicates the metatheory as the syntax of types is now
quotiented over the monoid laws and distributivity of composition over
choice. This complication makes type equivalence a decidable, but
difficult problem \cite{DBLP:conf/tacas/AlmeidaMV20} and subtyping an
undecidable problem
\cite{DBLP:journals/toplas/Padovani19,DBLP:journals/tcs/SilvaMV26}.

CFSTs are one starting point of our investigation. Symmetric
composition makes them well-suited to splitting local borrows. Only
remote borrows require extra synchronization machinery. Compared to
the CFST work and the implemented system FreeST based on it
\cite{almeida26:_frees_concur_progr_languag_with}, our calculus
does not consider polymorphism. We rely on the FreeST
implementation as an external constraint solver for splitting
constraints. 


The most closely related work is ``Borrowing from Session Types'' (BST) \cite{DBLP:journals/pacmpl/SaffrichS0V25}.
They present a similar approach to borrowing, but for regular session types, rather than CFST.
Our type system is heavily inspired by theirs, as it also relies on
BI-contexts that enforce a run-time ordering on operations.
We build on their translation from borrowing notation to explicit
splits, but adapt it to account for the distinction between local and
remote splits. The main technical differences are fourfold. First, BST
defines the semantics of borrowing by translation, whereas
\thecalculus has a direct operational semantics based on binder
groups. Second, \thecalculus distinguishes mobile, remote borrows from
bounded, local borrows (\cref{fig:bounded-session-types}); local
borrows are ordered by typing and evaluation and therefore avoid the
synchronization required by remote borrows. Third, splitting CFSTs
requires constraint generation modulo type equivalence rather than the
direct splitting algorithm available for traditional, tail-recursive
session types; we discharge these constraints with FreeST. Finally, we
define a low-level target calculus and prove a simulation result for
the translation, giving an implementation-oriented account of the
run-time synchronization protocol. The metatheory is mechanized in
Agda~\cite{Agda}.

Manifest sharing \cite{DBLP:journals/pacmpl/BalzerP17} is concerned with session types, which
make part of a protocol shareable. To access the shared part, they
employ a typed acquire-release pattern, which is reflected in the
types by different modalities. This approach makes the access pattern
manifest in the session type, similar to our pair of $\CDrop$-$\CAcq$
operators, but it may admit deadlock.  A subsequent work extends
manifest sharing to guarantee freedom of deadlock \cite{DBLP:conf/esop/BalzerTP19}.

Borrowing is not the only way to avoid resource-passing style in
session types. Several monadic
APIs for session types with a similar goal exist~\cite{DBLP:conf/haskell/PucellaT08,DBLP:journals/scp/ImaiYY19,DBLP:conf/padl/NeubauerT04,DBLP:journals/jfp/Padovani17}.
\citet{DBLP:journals/jfp/Padovani17} provides a monadic API on top of session
primitives that eliminates explicit channel manipulation for functions
that involve a single session. This API enables nesting of sessions,
but not interleaving multiple sessions. Its primitive operations can
``split'' a session type and their monadic
composition can split arbitrarily large prefixes.
Subsequent work
\cite{DBLP:conf/haskell/PucellaT08,sackman08:_session_types_haskel}
handles multiple sessions, but represents
them with a stack of capabilities at the type level. Monadic
composition splits prefixes of multiple sessions, but these prefixes are not
independent. \citet{DBLP:journals/corr/abs-1110-4163} extend this
work with session type inference and \citet{DBLP:journals/scp/ImaiYY19} provide an
update with further functionality.
In comparison, our proposal does not require type-level computations,
provides for interleaving multiple sessions, and for splitting
multiple session types independently at
arbitrary positions.

\subsection{Ordered Types}
\label{sec:ordered-types}

Ordered types go beyond linear types in eliding the exchange
principle. Their formal basis is  non-commutative logic
\cite{DBLP:journals/mlq/Abrusci90a, ABRUSCI199929} and they have been
introduced to a computer science setting by \citet{lambek58:_mathem_senten_struc,
  lambek60:_calcul_syntac_types}  to describe
sentence structure in natural languages.
Law and order \cite{DBLP:journals/pacmpl/Saffrich0024} is a precursor
of BST. It draws inspiration from work on
non-commutative logic and ordered lambda calculus
\cite{DBLP:conf/tlca/PolakowP99, DBLP:journals/entcs/PolakowP99}
and puts it to work in the context of typestate. It targets sequential
programming, so it is not concerned with synchronization issues, at all.

Other work on ordered types and lambda calculi is concerned with
memory management and memory layout
\cite{DBLP:conf/popl/PetersenHCP03} or stack allocation
\cite{DBLP:conf/tldi/AhmedW03}. Like Law and order, BST, and the
present work, these earlier works deviate quickly from the pure,
proof-theoretic form of ordered lambda calculus \cite{DBLP:conf/tlca/PolakowP99,
  DBLP:journals/entcs/PolakowP99}. Each develops a calculus that picks
up ideas, but puts significant domain-specific elements on top, to the
extent that even basic elimination principles
change. \citet{DBLP:conf/tldi/AhmedW03} share with our work the use of
BI-contexts to distinguish ordered and unordered binding groups.

\citet{DBLP:journals/pacmpl/CutlerWNHGSP24} present an application of
ordered types to typing a calculus for streaming data. Ordering in
their system expresses temporal ordering in an input stream, that is, a
stream of type $s\cdot t$ is a stream of type $s$ followed by a stream
of type $t$. They rely on the same tree-shaped context structure we use, but with a
different interpretation: their parallel context former behaves
intuitionistically; contrary to ours, it allows for weakening and
contraction. This system is very close to the original theory of
ordered types, but it is restricted to first-order functions. This
line of work has predecessors that consider streaming of
tree-structured data using ordered linear types
\cite{DBLP:journals/jfp/KodamaSK08,DBLP:journals/jip/SatoSK11},
specialized to a fixed binary tree representation.

The subtyping rule $(A;B) \| (C;D) \IsSSubtypeOf (A \| B) ; (C \| D)$
that weakens parallelism for ordered sequentiality seems to
originate from work on concurrent Kleene algebra
\cite{DBLP:conf/concur/HoareMSW09}, but \citet{CAIRES2008120} mentions it
earlier in a type system for concurrent computation. It also appears
in work on behavioral separation \cite{DBLP:conf/popl/CairesS13} where
it is attributed to \citet{DBLP:journals/tcs/BloomE96a}.

\subsection{Borrowing}
\label{sec:borrowing}

The notion of borrowing arose out of the need to control
aliasing in object oriented languages. It is closely tied to the notion of
ownership, which gave rise to numerous type disciplines to statically control
ownership. \citet{DBLP:series/lncs/ClarkeOSW13} present a survey of such
systems. Rust \cite{team24:_rust_progr_languag} is the most prominent
programming language supporting these concepts nowadays.
%
While ownership and borrowing are also useful concepts in the context
of session types, there are significant differences. Sessions are
resources for communication: using an endpoint advances its protocol
state. Thus there is no read-only analogue of an immutable borrowed
reference as it exists in Rust. Further, borrowing a session prefix
computes a residual session type, whereas Rust borrowing does not
temporally split ownership into a protocol prefix and suffix.

Other formalisms surrounding ownership and borrowing study
memory-like resources, including borrowing permissions
\cite{DBLP:conf/popl/NadenBAB12} and functional ownership systems
\cite{DBLP:journals/corr/abs-2310-18166}.
These systems address the interaction of linear ownership with
aliasing, permissions, and lifetimes. Our setting is different because
aliases are endpoints for communication protocols: their use changes
the protocol state and, once sent to another process, may require
run-time synchronization.

\citet{DBLP:journals/pacmpl/WagnerGML025} derive borrowing from a
linear lambda calculus for manual memory management. Their Borrow
Calculus supports immutable and mutable borrows, lexical lifetimes, and
reborrowing without adding borrowing primitives to the operational
semantics; soundness is semantic, via Borrow Logic, and yields
termination and leak freedom. We share the perspective that borrowing
is a disciplined way to relax linear resource passing temporarily.
However, our resources are session endpoints rather than memory
locations. This changes the problem: every borrowed endpoint is a
mutable communication capability, borrowing computes a prefix/suffix
decomposition of a session type, and remote borrows must be sequenced
by explicit synchronization. Thus our focus is on context-free session
type splitting, direct concurrent semantics, and an implementation-level
translation, rather than on a separation-logic model of memory
borrowing.

\citet{DBLP:journals/pacmpl/VillaniHDJ25} develop Tree Borrows, a
dynamic aliasing model for Rust that tracks derived references and raw
pointers in a tree of permissions. Like our target semantics, Tree
Borrows gives operational content to the idea that a borrow is a
derived capability whose use can restrict the original owner and other
aliases. The resemblance stops at this level of authority tracking:
Tree Borrows governs memory accesses and undefined behavior for Rust
pointers, whereas our borrows split session types into a borrowed
protocol prefix and a residual suffix. In particular, our remote
borrows require explicit synchronization to order communication on the
borrowed prefix before communication on the suffix; Tree Borrows has no
corresponding notion of protocol residual or cross-process handoff.

Overall, prior borrowing systems manage aliasing of memory-like
resources, whereas our contribution is to make borrowing compute and
schedule portions of communication protocols.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../popl2027.tex"
%%% End:
