\section{Dynamics}
\label{sec:semantics}

The dynamics is a small-step semantics for the core language of
\cref{fig:syntax}.  Expressions reduce inside the evaluation contexts shown
below; processes reduce by lifting expression steps to threads and by
synchronizing the communication and borrowing operations that act on binder
groups.

\begin{align*}
  \U,\V \grmeq
    & \C
      \grmor \EVar
      \grmor \EAbs \EVar\E
      \grmor \EInj{i} \V
      \grmor \EPair \U\V
      % \grmor \CSend \V 
              \tag{Values}
  \\
  \EC \grmeq&
              \ECEmpty
              \grmor \EAppD \EC\E\DirLeft
              \grmor \EAppD \V\EC\DirLeft
              \grmor \EAppD \E\EC\DirRight
              \grmor \EAppD \EC\V\DirRight
              \grmor \ESeq \EC\E
              \grmor \EPair \EC\E
              \grmor \EPair \V\EC
              \\ \grmor
    & \EInj{i}\EC \grmor \ELet {\EPair \EVar\EVar}\EC\E
              \grmor \ECase \EC {\ECaseBranch* i \EVar \E}
              \tag{Evaluation contexts}
  \\
  \FC \grmeq
  & \PExp \EC
    \tag{Thread evaluation contexts}
  \\
  \CC \grmeq
  & \ECEmpty
    \grmor \PPar{\CC}{\Proc}
    \grmor \PNew \BindGroup\BindGroup \CC
    \tag{Process contexts}
\end{align*}

Values are the usual call-by-value normal forms: constants, variables,
abstractions, injections of values, and pairs of values.  The expression
contexts fix the deterministic order of evaluation, including the order chosen
by a directed application.  A thread context embeds an expression context in a
single running thread, whereas process contexts describe the positions closed
under the process congruence and closure rules below.  The expression reduction
relation, given in \cref{fig:expression-reduction}, contains the standard
computational steps for beta-reduction, sequencing, pair and variant
elimination, recursive unfolding, and closure under expression contexts.

\Cref{fig:process-reduction} gives the process reduction relation.  Its rules
fall into three groups: ordinary computation inside a thread, synchronization
between the two sides of a restricted channel, and structural closure.

\begin{figure}[t!]
  \begin{mathpar}
    \RuleProcRedExp \and
    \RuleProcRedNew \and
    \RuleProcRedFork \and
    \RuleProcRedCom \and
    \RuleProcRedLSplit \and
    \RuleProcRedRSplit \and
    \RuleProcRedDiscard \and
    \RuleProcRedDrop \and
    \RuleProcRedAcquire \and
    \RuleProcRedClose \and
    \RuleProcRedChoice \and
    \RuleProcRedPar \and
    \RuleProcRedBind \and
    \RuleProcRedStruct
  \end{mathpar}
  \caption{Process reduction ($\Proc \PReduce \Proc$)}
  \label{fig:process-reduction}
\end{figure}

Rule \textsc{R-Exp} embeds an expression step in the selected thread context.
Rule \textsc{R-New} allocates a fresh restricted channel and returns the two
fresh endpoints as a pair.  Each endpoint is placed behind an initial separator
in its binder group, matching the acquired session type assigned to channel
creation in \cref{fig:process-typing}.  Rule \textsc{R-Fork} spawns the
thunked process in parallel and returns unit to the forking thread.

Communication rules operate on the heads of dual binder groups.  Rule
\textsc{R-Com} synchronizes send and receive, transfers the transmitted value,
and advances both groups.  Rule \textsc{R-Choice} similarly synchronizes
selection and branching, returning the selected label as a variant.  Rule
\textsc{R-Close} synchronizes close and wait at the end of a session, and rule
\textsc{R-Discard} consumes an endpoint whose type permits it to be skipped.

The split rules implement the local and remote structure described in
\cref{sec:direct-semant-chann}.  Rule \textsc{R-LSplit} replaces one endpoint
name by two ordered names in the same binder group and returns the ordered
pair.  Because a binder group is a sequence, subsequent reductions must consume
the left component before the right component.  Rule \textsc{R-RSplit} replaces
one endpoint name by two names separated by a binder-group separator and
returns them as an unordered pair.  The separator is the runtime marker for the
remote handoff tracked statically by the $\SDrop$ and $\SAcq$ constructors in
\cref{fig:process-typing}.

Rules \textsc{R-Drop} and \textsc{R-Acquire} move across such a separator.
Dropping removes the endpoint immediately to the left of the separator and
exposes the separator at the front of the remaining group.  Acquiring can fire
only once the separator is exposed; it removes the separator and returns the
endpoint on the right.  Thus remote splitting transfers control from one
thread to another only after the prefix of the original endpoint has been
consumed.

Finally, \textsc{R-Par} and \textsc{R-Bind} close reduction under parallel
composition and restriction.  Rule \textsc{R-Struct} closes the relation under
the process congruence of \cref{fig:configuration-congruence}, so reductions
may use associativity, commutativity, and scope movement to expose a redex.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../popl2027.tex"
%%% End:
